Commutative Character Sheaves

### 2015-11-15 followup.tex
Added objects and automorphism groups to Section 3.

Next, we broaden our scope further to encompass smooth group schemes $G$ that are not necessarily commutative.
In this paper, we do not consider Lusztig's geometrization of arbitrary characters of the non-abelian group $G(k)$
using character sheaves \cite{lusztig:85a}*{Def.~2.10}, but focus only on the case of $1$-dimensional characters
$\Hom(G(k), \EEx)$.  The category $\CS(G)$ has a straightforward generalization to this case, but again
there are more linear character sheaves than there are characters, as pointed out by Kamgarpour \cite{kamgarpour:09a}*{(1.1)}.
We define a category $\CCS(G)$ and a forgetful functor to $\CS(G)$ so that $\TrFrob{G} : \CCSiso{G} \to \Hom(G\ab(k), \EEx)$
is an isomorphism.  In fact, pullback along the quotient $q : G \to G\ab$ defines an equivalence of categories
$\CCS(G\ab) \to \CCS(G)$.  Note that the functor $\CCS(G) \to \CS(G)$ is not essentially surjective, missing
the kinds of linear character sheaves highlighted by Kamgarpour.  Moreover, $\Hom(G\ab(k), \EEx)$ surjects
onto $\Hom(G(k), \EEx)$, with a kernel described by Galois cohomology.

@@

Every $\beta : \cs{L}\vert_{G\der} \to (\EE)_{G\der}$ determines an isomorphism $\gamma : m^*\cs{L} \to \theta^*m^*\cs{L}$ as follows.
Let $c : G\times G\to G\der$ be the commutator map, defined by $c(x,y)= xyx^{-1}y^{-1}$.
Then $c$ is a smooth morphism of $\Fq$ schemes.
Set $m' = i \circ m \circ \theta$; then $j\circ c = m \circ (m \times m')$.
Thus, $\beta : \cs{L}\vert_{G\der} \to (\EE)_{G\der}$ determines the isomorphism $\gamma' : m^*\cs{L} \otimes \theta^* m^* i^*\cs{L} \to (\EE)_{G\times G}$ by the diagram of isomorphisms, below.
\begin{equation}
\begin{tikzcd}
\arrow[equal]{d} c^* (\cs{L}\vert_{G\der}) \arrow{r}{c^*(\beta)} 
	&  c^*((\EE)_{G\der}) \arrow[equal]{d} \\
\arrow{d}[swap]{j\circ c = m \circ (m \times m')} c^* j^* \cs{L} 
	&   (\EE)_{G\times G} \\
(m \times m')^* m^* \cs{L} \arrow{d}[swap]{(m \times m')^*(\mu)} 
	&  m^*\cs{L} \otimes \theta^* m^* i^* \cs{L}  \arrow[dashed]{u}[swap]{\gamma'} \\
(m \times m')^* (\cs{L} \boxtimes \cs{L}) \arrow[equal]{r} 
	& m^*\cs{L} \otimes (m')^*\cs{L} \arrow{u}[swap]{m' = i\circ m\circ \theta} 
\end{tikzcd}
\end{equation}
In the diagram above, the arrows labeled with equations come from canonical isomorphisms of functors on Weil sheaves derived from the equations; so, for example, the middle left isomorphism comes from $(m\times m')^* m^* \iso c^* j^*$ since $j\circ c = m \circ (m \times m')$.
Using the monoidal structure of the category of Weil local systems on $G\times G$, this isomorphism, in turn, defines an isomorphism
$
m^*\cs{L} \to (\theta^* m^* i^*\cs{L})^\vee.
$
Since $(\theta^* m^* i^*\cs{L})^\vee \iso \theta^* m^* i^* (\cs{L}^\vee)$ canonically, and since $\cs{L}^\vee \iso i^*\cs{L}$, canonically, this defines the promised isomorphism
\[
\begin{tikzcd}
\gamma : m^*\cs{L} \arrow{r} & \theta^* m^* \cs{L}.
\end{tikzcd}
\]

@@

\todo{We should give the relationship between $\CCSiso{G}$ and $G(k)^*$ here as a corollary.  Also, include the automorphism group of a commutative character sheaf.}

\todo{Comment here on Masoud's work.}

@@

We may now revisit the definition of the functor $q_H^*: \CS(G_0)\to \CS_H(G)$. 
On objects, $q_H^*: \CS(Q)\to \CS_H(G)$ is given by $q_H^*(\cs{L}_0,\mu_0) = (L_1(\cs{L}_0), L_2(\mu_0))$. \todo{Shorten this, actually, and combine with the next paragraph.}
To verify that $(L_1(\cs{L}_0), L_2(\mu_0))$ is an object of $\CS_H(G)$, one must confirm \cite{cunningham-roe:13a}*{CS1, CS2, CS3}, \eqref{E1}, \eqref{E2} and \eqref{ECS1}. 
Conditions \cite{cunningham-roe:13a}*{CS1, CS2, CS3} follow from an adaptation of \cite{cunningham-roe:13a}*{} while conditions \eqref{E1}, \eqref{E2} are a consequence of the work above. 
Condition \eqref{ECS1} is confirmed by brute force.
On maps, $q_H^*: \CS(Q)\to \CS_H(G)$ is given by $q_H^*(\alpha_0) = L_1(\alpha_0)$.
To verify that $L_1(\alpha_0)$ is a map in $\CS_H(G)$ we must confirm conditions \cite{cunningham-roe:13a}*{CS4} and \eqref{E3}.
Condition \cite{cunningham-roe:13a}*{CS4} follows from an adaptation of \cite{cunningham-roe:13a}*{} while condition \eqref{E3} is a consequence of the work above.

With this description of $q_H^*: \CS(Q)\to \CS_H(G)$ is is easy to see that it is an equivalence.
Its adjoint is given on objects by $(\cs{L},\mu,\nu) \mapsto (R_1(\cs{L},\nu), R_2(\nu))$, making use of Lemma~\ref{lem:HH}, and on morphisms by $\alpha \mapsto R_1(\alpha)$.
The adjunction is built from the adjunctions $(\epsilon_1, \eta_1)$ and $(\epsilon_2, \eta_2)$ for $(L_1,R_1)$ and $(L_2, R_2)$.
\end{proof}

@@

\subsection{Commutative quasicharacter sheaves}

>>>>

Next, we broaden our scope further to encompass smooth group schemes $G$ that are not necessarily commutative.
In this paper, we do not consider Lusztig's geometrization of arbitrary characters of the non-abelian group $G(k)$, when $G$ is reductive,
using character sheaves \cite{lusztig:85a}*{Def.~2.10}, but focus on the case of $1$-dimensional characters
$\Hom(G(k), \EEx)$ assuming only that $G$ is smooth. The category $\CS(G)$ has a straightforward generalization to this case, but again
there are more linear character sheaves than there are characters, as pointed out by Kamgarpour \cite{kamgarpour:09a}*{(1.1)}.
We define a category $\CCS(G)$ and a forgetful functor to $\CS(G)$ so that $\TrFrob{G} : \CCSiso{G} \to \Hom(G\ab(k), \EEx)$
is an isomorphism.  In fact, pullback along the quotient $q : G \to G\ab$ defines an equivalence of categories
$\CCS(G\ab) \to \CCS(G)$.  Note that the functor $\CCS(G) \to \CS(G)$ is not essentially surjective, missing
the kinds of linear character sheaves highlighted by Kamgarpour.  Moreover, $\Hom(G\ab(k), \EEx)$ surjects
onto $\Hom(G(k), \EEx)$, with a kernel described by Galois cohomology.

@@

Every $\beta : \cs{L}\vert_{G\der} \to (\EE)_{G\der}$ determines an isomorphism $\gamma : m^*\cs{L} \to \theta^*m^*\cs{L}$ as follows.
Let $c : G\times G\to G\der$ be the commutator map, defined by $c(x,y)= xyx^{-1}y^{-1}$.
Then $c$ is a smooth morphism of $\Fq$ schemes.
Set $m' = i \circ m \circ \theta$; then $j\circ c = m \circ (m \times m')$.
Then, $\beta : \cs{L}\vert_{G\der} \to (\EE)_{G\der}$ determines the isomorphism $\gamma' : m^*\cs{L} \otimes \theta^* m^* i^*\cs{L} \to (\EE)_{G\times G}$ by the diagram of isomorphisms, below.
\begin{equation}
\begin{tikzcd}
\arrow[equal]{d} c^* (\cs{L}\vert_{G\der}) \arrow{r}{c^*(\beta)} 
	&  c^*((\EE)_{G\der}) \arrow[equal]{d} \\
\arrow{d}[swap]{j\circ c = m \circ (m \times m')} c^* j^* \cs{L} 
	&   (\EE)_{G\times G} \\
(m \times m')^* m^* \cs{L} \arrow{d}[swap]{(m \times m')^*(\mu)} 
	&  m^*\cs{L} \otimes \theta^* m^* i^* \cs{L}  \arrow[dashed]{u}[swap]{\gamma'} \\
(m \times m')^* (\cs{L} \boxtimes \cs{L}) \arrow[equal]{r} 
	& m^*\cs{L} \otimes (m')^*\cs{L} \arrow{u}[swap]{m' = i\circ m\circ \theta} 
\end{tikzcd}
\end{equation}
In the diagram above, the arrows labeled with equations come from canonical isomorphisms of functors on Weil sheaves derived from the equations; so, for example, the middle left isomorphism comes from $(m\times m')^* m^* \iso c^* j^*$ since $j\circ c = m \circ (m \times m')$.
Using the monoidal structure of the category of Weil local systems on $G\times G$, the isomorphism $\gamma' : m^*\cs{L} \otimes \theta^* m^* i^*\cs{L} \to (\EE)_{G\times G}$ defines an isomorphism
\[
m^*\cs{L} \to (\theta^* m^* i^*\cs{L})^\vee.
\]
Since $(\theta^* m^* i^*\cs{L})^\vee \iso \theta^* m^* i^* (\cs{L}^\vee)$ canonically, and since $\cs{L}^\vee \iso i^*\cs{L}$, canonically, this defines the promised isomorphism
\[
\begin{tikzcd}
\gamma : m^*\cs{L} \arrow{r} & \theta^* m^* \cs{L}.
\end{tikzcd}
\]

@@

\subsection{Objects and maps in commutative character sheaves}

Theorem~\ref{thm:Gab} shows that $\CCS(G)$ is a categorical solution to the problem that linear character sheaves on $G$ need not be trivial on $G\der$, as discussed at the beginning of Section~\ref{sec:noncom}; at the same time, it is the categorical solution to the problem, discussed in Section~\ref{ssec:defect}, that the dictionary between linear character sheaves on $G\ab$ and characters of $G\ab$, need not be perfect.
%
But a description of $\CCS(G)$ requires, at the very least, a description of isomorphism classes of objects, and maps. 
%

\todo{We should give the relationship between $\CCSiso{G}$ and $G(k)^*$ here as a corollary.  Also, include the automorphism group of a commutative character sheaf.}
\todo{Is this what you had in mind?}

\begin{proposition}
The trace of Frobenius $\TrFrob{} : \CCSiso{G}\to G(\Fq)^*$ fits into the following diagram,
\[
\begin{tikzcd}
\ & & \CCSiso{G\ab} \arrow{d}{\TrFrob{}}[swap]{\iso} \arrow{r}{\iso} & \CCSiso{G} \arrow{d}{\TrFrob{}}& \\
1 \arrow{r} & \Delta^* \arrow{r} & G\ab(k)^* \arrow{r} & G(k)^* \arrow{r} & 1,
\end{tikzcd}
\]
where $\Delta$ denotes the image of the connecting homomorphism $G\ab(\Fq) \to \Hh^1(\Fq, G\der)$.
In particular, isomorphism classes of objects in $\CCS(G)$ are canonically identified with the group $G\ab(\Fq)^*$.
\end{proposition}

\begin{proof}
Let $\Delta$ be the image of the boundary homomorphism $\delta$ in the sequence
\[
1 \to G\der(k) \to G(k) \to G\ab(k) \xrightarrow{\delta} \Hh^1(k, G\der),
\]
yielding a short exact sequence
\[
1 \to G(k) / G\der(k) \to G\ab(k) \to \Delta \to 1.
\]
Applying $\Hom(-, \EEx)$ and using the fact that every homomorphism $G(k) \to \EEx$ vanishes on $G\der(k)$, we get
\[
1 \to \Delta^* \to G\ab(k)^* \to G(k)^* \to 1.
\]
By Theorem~\ref{thm:Gab}, the trace of Frobenius map 
\end{proof}

\begin{proposition}
Every map in $\CCS(G)$ is either trivial or an isomorphism. 
For every commutative character sheaf on $G$, its automorphism group is
$
\Hh^1(\pi_0(G\ab),\EEx)^{\Frob{}}.
$
\end{proposition}

\begin{proof}
$= \Aut_{\CCS(G\ab)}(\cs{L}\ab,\mu\ab) =$
\end{proof}

@@

We may now revisit the definition of the functor $q_H^*: \CS(G_0)\to \CS_H(G)$:
on objects, $q_H^*: \CS(Q)\to \CS_H(G)$ is given by $q_H^*(\cs{L}_0,\mu_0) = (L_1(\cs{L}_0), L_2(\mu_0))$; on maps, $q_H^*: \CS(Q)\to \CS_H(G)$ is given by $q_H^*(\alpha_0) = L_1(\alpha_0)$.
Direct calculation confirms \cite{cunningham-roe:13a}*{CS1, CS2, CS3}, \eqref{E1}, \eqref{E2} and \eqref{ECS1}, and therefore that $(L_1(\cs{L}_0), L_2(\mu_0))$ is an object of $\CS_H(G)$; 
likewise, $L_1(\alpha_0)$ is a map in $\CS_H(G)$ after checking \cite{cunningham-roe:13a}*{CS4} and \eqref{E3}.
%
With this description of $q_H^*: \CS(Q)\to \CS_H(G)$ is is easy to see that it is an equivalence.
Its adjoint is given on objects by $(\cs{L},\mu,\nu) \mapsto (R_1(\cs{L},\nu), R_2(\nu))$, making use of Lemma~\ref{lem:HH}, and on morphisms by $\alpha \mapsto R_1(\alpha)$.
The adjunction is built from the adjunctions $(\epsilon_1, \eta_1)$ and $(\epsilon_2, \eta_2)$ for $(L_1,R_1)$ and $(L_2, R_2)$.
\end{proof}

@@

\subsection{Linear characters of certain $p$-adic groups}
